Sketch Recognition using Particle Swarm Optimizations

### 2009-05-27 trunk/paper2/Paper2.tex
change to the segmentation and modification

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}

@@

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DB), Electrical Design (EL-DB) and Logic Design (LD-DB) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}

\section{Introduction}

@@

 After each iteration the global best $g_{best}$ particle and the agent local best particle $l_{best}$are evaluated based on the maximum fitness functions of all particles in the solution space. The solution is found after achieving a specific number of iteration or after an error threshold is achieved.
Equation \ref{eq:descrite} is used to change the general swarm algorithm into binary particle (\textit{Discrete Particle Swarm Algorithm DPSO}). The \textit{DPSO} handles particle values of either $0$ or $1$ \cite{PSODisceret}.  
\begin{equation}
   P(i)\Leftarrow 
\{
\begin{array}{c} 
1 \quad \quad if\quad r_{3}>p_{i}  \\

0 \quad \quad if\quad r_{3}<p_{i}, 
\label{eq:descrite}
\end{array}\}
\end{equation}
 where $p_{ij}$ is the numerical values of the particle and $r_{3}$ is a random variable.

@@

 \begin{equation}
%eff= (P_{percent}/E)
%eff = \left\{ {\begin{array}{*{20}c}
%   {\frac{{P_{percent} }}{E}} & {ifP_{percent}  < 0.5}  \\
%   {\frac{{P_{percent} }}{{E \times 2}}} & {ifP_{percent}  \ge 0.5}  \\
%\end{array}} \right,
eff= \begin{cases} 
\frac{P_{percent}}{E},&if P_{percent}< 0.5 \\
\frac{P_{percent}}{2\times E},&if P_{percent}\geq 0.5 \\
% \frac{P_{percent}}{2\times E},&\mbox{if} P_{percent}\mbox{\geq0.5} 
\end{cases},
\label{eq:circleError}
\end{equation}
 where $E$ is the error computed by Equation (\ref{eq:circleFit}), and 
\[
P_{percent}  = L_{stroke} /P_{ellipse}, 
%\label{eq:ErrorArea}
\]
 where $L_{stroke}$ total length of stroke and $P_{ellipse} $ is the perimeter of the estimated ellipse. If $eff$ has a value more than threshold $th_{Ellipse}$\footnote{By trial and error best threshold found was $th_{Ellipse}=0.2$} then the stroke is segmented as an ellipse otherwise the system proceeds to the next step.

@@

	\item[Fitness function:] The fitness function and error calculation are different in each of the two \textit{DPSO} algorithms. 
	\end{description}
\subsubsection{\textit{Polygon Approximation Algorithm \textsl{(AlgS1)}}}
%\\
The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The graphical meaning of the error is shown in Figure\ref{fig:LabelsPSO}.
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation} where the arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point $x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$. The line $\overline{x_i x_j}$ is defined as the straight line connecting point $x_i$ to point $x_j$ and $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$. The fitness is computed using the following equation %\ref{eq:fitnessSwarm1}
\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   {\frac{-E}{\varepsilon N}} & {ifE > \varepsilon ,}  \\

   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation} where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that was previously labeled as a possible dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a -ve value to lower the fitness value of the solution. Otherwise the system favors the lower number of vertices.

 \subsubsection{\textit{Hybrid Fitting \textsl{AlgS2}}}
The algorithm has the same problem formulation but different fitness and error functions are used. An attempt is made to fit each segment into a line or circular arc. The errors of both circle and line fit estimations are computed for each segment $S_i=\widehat{x_ix_j}$. The approximation with the lower error value is chosen as the final approximation of this segment $S_i$\cite{CruveDivisionSwarm}. The sum of the approximation errors of all segments is the total error of the particle.  The total error of the particle is computed by equation %\ref{eq:errorSwarm2}
 \begin{equation}
E=\sum\nolimits_{i = 0}^M e(D_i) 
\label{eq:errorSwarm2}
\end{equation}where $M$ is the number of segments in the solution as generated by the swarm algorithm, $D_i$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ where $d_c$ is the circle approximation error and $d_l$ is the line approximation error as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation %\ref{eq:fitnessSwarm2} 
\begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is the number of segments and $k$ is a parameter tweaked to get minimum number of segments. The larger $k$ is, the more effect the number of segments will have. For our system, $k$ is selected to be 0.5\cite{CruveDivisionSwarm}.

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, one of these points are removed based on the error cased by its removal.
	\item Each two successive segments are test to determine if they can be merged into a single segment. If merging the two segments have the same or smaller the error, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features(FS1)}
  Features defines the structure of the geometrical symbol.  
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} Number of segments in the symbol.
	 \item \emph{Strokes:} Number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item \emph{Curves:} Number of curves or ellipses in the symbol. 
		\item \emph{Lines:} Number of lines in the symbol. 
		\item \emph{Perpendicular} lines Number of perpendicular lines. 
		\item \emph{Parallel} lines Number of parallel lines. 
		\item \emph{Intersections} Number of intersection between lines and curves. 
		\item \emph{T intersections} Number of T intersections. 
		\item \emph{L intersections} Number of L intersections. 
		\item \emph{X intersections} Number of X intersections.

>>>>

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{array}
\setlength{\extrarowheight}{0.3cm}

@@

%\begin{abstract}
%%This study 
%Sketches and drawings are widely used as a simple method of expressing thought and designs. The goal of this paper is developing a sketch recognition system that facilitates using sketches in computer systems. The system uses Particle Swarm Optimization (PSO) algorithm to optimally segment the strokes the user draws into meaningful geometric primitives.  These geometric primitives are grouped to formulate symbols which are further identified using a Support Vector Machines (SVM) classifier. Two different segmentation algorithms were evaluated and results show that both algorithms improve the final recognition results. Experiments were conducted on three different dataset, simple presentation (Hs-DB), Electrical Design (EL-DB) and Logic Design (LD-DB) symbols. Results show that the effectiveness of the proposed system on various dataset especially on datasets with low shape diversity as in Logic Design dataset (LD-DS).  % sketch are divided  Users draw symbols and sketchs using a set of pen strokes. 
%%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple %strokes. Usually, users draw strokes using pens. The sketch recognition system immediately interprets their %strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization (PSO) to divide %the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to %formulate symbols which are further identified. The results show that using PSO improves segmentation results %which guide the symbol recognition phase. As for recognition we use Support Vector Machines (SVM) classifier %which further improves the final recognition accuracy.  
%\end{abstract}
\pagenumbering{arabic}
\section{Introduction}

@@

 After each iteration the global best $g_{best}$ particle and the agent local best particle $l_{best}$are evaluated based on the maximum fitness functions of all particles in the solution space. The solution is found after achieving a specific number of iteration or after an error threshold is achieved.
Equation \ref{eq:descrite} is used to change the general swarm algorithm into binary particle (\textit{Discrete Particle Swarm Algorithm DPSO}). The \textit{DPSO} handles particle values of either $0$ or $1$ \cite{PSODisceret}.  
\begin{equation}
   P(i)\Leftarrow 
\left \{
\begin{array}{c} 
1 \quad \quad if\quad r_{3}>p_{i}  \\

0 \quad \quad if\quad r_{3}<p_{i}, 
\label{eq:descrite}
\end{array}\right.
\end{equation}
 where $p_{ij}$ is the numerical values of the particle and $r_{3}$ is a random variable.

@@

 \begin{equation}
%eff= (P_{percent}/E)
%eff = \left\{ {\begin{array}{*{20}c}
%   {\frac{{P_{percent} }}{E}} & {ifP_{percent}  < 0.5}  \\
%   {\frac{{P_{percent} }}{{E \times 2}}} & {ifP_{percent}  \ge 0.5}  \\
%\end{array}} \right,
eff= \begin{cases} 
\displaystyle \frac{P_{percent}}{E},&if P_{percent}< 0.5 \\ \medskip
\displaystyle  \frac{P_{percent}}{2\times E},&if P_{percent}\geq 0.5 \\
% \frac{P_{percent}}{2\times E},&\mbox{if} P_{percent}\mbox{\geq0.5} 
\end{cases},
\label{eq:circleError}
\end{equation}
 where $E$ is the error computed by Equation (\ref{eq:circleFit}), and 
\[
P_{percent}  = L_{stroke} /P_{ellipse}, 
%\label{eq:ErrorArea}
\]
 where $L_{stroke}$ total length of stroke and $P_{ellipse} $ is the perimeter of the estimated ellipse. If $eff$ has a value more than threshold $th_{Ellipse}$\footnote{By trial and error best threshold found was $th_{Ellipse}=0.2$} then the stroke is segmented as an ellipse otherwise the system proceeds to the next step.

@@

	\item[Fitness function:] The fitness function and error calculation are different in each of the two \textit{DPSO} algorithms. 
	\end{description}
\subsubsection{\textit{Polygon Approximation Algorithm \textsl{(AlgS1)}}}
%\\
The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The graphical meaning of the error is shown in Figure\ref{fig:pso1}.
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation} where the arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point $x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$. The line $\overline{x_i x_j}$ is defined as the straight line connecting point $x_i$ to point $x_j$ and $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$. The fitness is computed using the following equation %\ref{eq:fitnessSwarm1}

\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{30}l}
   {\displaystyle \frac{-E}{\varepsilon N}} & {ifE > \varepsilon }  \\

   { \frac {D}{\sum\limits_{j = 1}^N {p_{ij} } }} & {otherwise}  \\
\end{array}} \right.,
\label{eq:fitnessSwarm1}
\end{equation} where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that were previously labeled as possible dominant points ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a $-ve$ value to lower the fitness value of the solution. Otherwise, the system favors the least number of vertices.

 \subsubsection{\textit{Hybrid Fitting Algorithm \textsl{(AlgS2)}}}
The \textsl{(AlgS2)} algorithm has the same problem formulation as \textsl{(AlgS1)}. However, it has different fitness and error functions. An attempt is made to fit each segment into a line or circular arc. The errors of both circle and line fit estimations are computed for each segment $S_i=\widehat{x_ix_j}$. The approximation with the lowest error value is chosen as the final approximation of this segment $S_i$\cite{CruveDivisionSwarm}. The sum of the approximation errors of all segments is considered the total error of the particle. The total error of the particle is computed by equation %\ref{eq:errorSwarm2}
 \begin{equation}
E=\sum\nolimits_{i = 0}^{N_s} e(D_i) 
\label{eq:errorSwarm2}
\end{equation}where $N_s$ is the number of segments in the solution as generated by the swarm algorithm, $D_i$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ where $d_c$ is the circle approximation error and $d_l$ is the line approximation error as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation %\ref{eq:fitnessSwarm2} 
\begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times {N_s}^k }}
\label{eq:fitnessSwarm2}%$N_s$ is the number of segments and
\end{equation} where $E$ is the error and $k$ is a parameter tweaked to get minimum number of segments. The larger $k$ is, the higher the effect the number of segments will have. For our system, $k$ is selected to be 0.5 \cite{CruveDivisionSwarm}.

After each iteration of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedures: 
\begin{enumerate}
	\item For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{Prepross}). If it was not labeled, the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. 
	\item The particles are tested to make sure that the distance between every two successive dominate point is larger than $min_D$, where $min_D$ is 5\% of the total length of the stroke.  Otherwise, one of these points is removed based on the error cased by its removal.
	\item Each two successive segments are test to determine if they can be merged into a single segment. If merging the two segments have the same or smaller the error, the two segments are merged. This step ensures that the final segmentation of the system has minimum number of segments. 
\end{enumerate}
%1) 2)  % If two points are nearer than $min_D$ then one of the points is removed. 
\subsection{Recognition}
\label{sec:Recognition}
After the user draws all strokes of the symbol, the set of un-recognized strokes is grouped together along with their segmentation as input to the feature extraction process. A composite set of features is used to generate a single feature vector. The following list describes each feature in details.
%\begin{description}
\subsubsection{Structural and geometrical Features(FS1)}
  Features defines the structure of the geometrical symbol.  
%\begin{description}
%	 \item[No. Of segments] Number of segments in the symbol. 
%		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
%		\item [No. of curves] Number of curves or ellipses in the symbol. 
%		\item [No. of lines] Number of lines in the symbol. 
%		\item [No. of perpendicular lines] Number of perpendicular lines. 
%		\item [No. of parallel lines] Number of parallel lines. 
%		\item [No. of intersections] Number of intersection between lines and curves. 
%		\item [Size Ratio] Ratio between width to height of the symbol.
%		%\item[No. of holes] Number of holes in the symbols. 
%
%\end{description}
 \begin{itemize}
	 \item \emph{Segments:} Number of segments in the symbol.
	 \item \emph{Strokes:} Number of strokes or partial strokes that created the symbols.  
		\item  \emph{Primitives:} Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item \emph{Curves:} Number of curves or ellipses in the symbol. 
		\item \emph{Lines:} Number of lines in the symbol. 
		\item \emph{Perpendicular} lines Number of perpendicular lines. 
		\item \emph{Parallel} lines Number of parallel lines. 
		\item \emph{Intersections} Number of intersection between lines and curves. 
		\item \emph{T intersections} Number of T intersections. 
		\item \emph{L intersections} Number of L intersections. 
		\item \emph{X intersections} Number of X intersections.
